Vragenlijst thesis evaluatie

### 2015-08-19 doc/template/chap-4.tex
veranderingen opslagen
finaal

\section{Methods of propagation}
\label{methodsofpropagation}

@@

\section{Models for worm propagation}
\label{modelsforpropagation}
Early work on worm propagation models are based on the spread of real-world epidemic diseases. These models are based on the transition state of each node in the network.\\

@@

\section{Matrix worm model}
\label{eigenmatrixmethode}

@@

%What do we need for an algorithm
%\begin{description}
%\item Graph network $G = < V, E>$
%\item Graph matrix $[A]$ which is $|V| \times |V| $
%\item Attack vector $[X]$ which is $1 \times |V|$
%\item cummulative matrix $[M]$ which is $|V| \times |V|$
%\item state matrix $[T]$  which is $|V| \times |V|$
%\item Reset vector $[R]$
%\item duration \textit{d}
%\item time \textit{n}
%\item rate $\delta _{0}$ of defender and $\delta _{1}$ of attacker
%\end{description}
%
%
%
%Initialisation algorithm:
%
%
%\begin{verbatim}
%initialisatie
%	d=0
%	A=basismatrix
%	M=A^{0}
%	n=0
%	\delta_{0}
%	\delta_{1}
%	X
%	R
%	controller = defender
%	
%	
%
%	Algorithm
%	n:= n + 1;
%	Check who is in control? ( through modulo )
%	if ( defender & controller=defender)
%				d:= d + 1;
%	
%	if ( defender & controller=attacker )
%				G = X \times R  (flippen ten voordele van defender)
%				d = 0
%				controller = defender
%				
%	if ( attacker & controller=defender )
%				controller=attacker
%				..
%				
%	if ( attacker & contoller=attacker )
%				d:= d + 1
%				M = M x A
%				T = T + M
%				G = X x T
%				
%		
%\end{verbatim}
\section{PageRank algorithm}
\label{googlesec:5.4}

@@

The PageRank of each page on the internet is the dominant eigenvalue of the Google Matrix. The Google Matrix is a stochastic matrix and represents a graph where every edge denotes a link between pages. Through the power iteration  \ref{Arnoldi}, the PageRank of all the pages can be calculated iteratively. \\

@@

We can use the Google matrix to calculate the probability that a node in a graph is infected by a worm after a certain amount of time. When a worm uses the topology of the graph to propagate the damping variable p must be set equal to zero, as in the worm will not suddenly attack a node that is not connected. If the propagation of a worm is topology independent, the damping variable \textit{p} can be set to 1. \\
The ranking of nodes provides interesting avenues for further research for the case where the defender flips a subset of nodes (rather than the entire network). It can be advantageous to flip the nodes with a higher PageRank more often than the other nodes. A higher PageRank means that the page has a higher probability of being compromised by a worm.

>>>>

\section{Methods of Propagation}
\label{methodsofpropagation}

@@

\section{Models for Worm Propagation}
\label{modelsforpropagation}
Early work on worm propagation models are based on the spread of real-world epidemic diseases. These models are based on the transition state of each node in the network.\\

@@

\section{Matrix based Worm Model}
\label{eigenmatrixmethode}

@@

%What do we need for an algorithm
%\begin{description}
%\item Graph network $G = < V, E>$
%\item Graph matrix $[A]$ which is $|V| \times |V| $
%\item Attack vector $[X]$ which is $1 \times |V|$
%\item cummulative matrix $[M]$ which is $|V| \times |V|$
%\item state matrix $[T]$  which is $|V| \times |V|$
%\item Reset vector $[R]$
%\item duration \textit{d}
%\item time \textit{n}
%\item rate $\delta _{0}$ of defender and $\delta _{1}$ of attacker
%\end{description}
%
%
%
%Initialisation algorithm:
%
%
%\begin{verbatim}
%initialisatie
%	d=0
%	A=basismatrix
%	M=A^{0}
%	n=0
%	\delta_{0}
%	\delta_{1}
%	X
%	R
%	controller = defender
%	
%	
%
%	Algorithm
%	n:= n + 1;
%	Check who is in control? ( through modulo )
%	if ( defender & controller=defender)
%				d:= d + 1;
%	
%	if ( defender & controller=attacker )
%				G = X \times R  (flippen ten voordele van defender)
%				d = 0
%				controller = defender
%				
%	if ( attacker & controller=defender )
%				controller=attacker
%				..
%				
%	if ( attacker & contoller=attacker )
%				d:= d + 1
%				M = M x A
%				T = T + M
%				G = X x T
%				
%		
%\end{verbatim}
\section{PageRank Algorithm}
\label{googlesec:5.4}

@@

The PageRank of each page on the internet is the dominant eigenvalue of the Google Matrix. The Google Matrix is a stochastic matrix and represents a graph where every edge denotes a link between pages. Through the power iteration \citep{Arnoldi}, the PageRank of all the pages can be calculated iteratively. \\

@@

We can use the Google matrix to calculate the probability that a node in a graph is infected by a worm after a certain amount of time. When a worm uses the topology of the graph to propagate the damping variable p must be set equal to zero, as in the worm will not suddenly attack a node that is not connected. If the propagation of a worm is topology independent, the damping variable \textit{p} can be set to 1. \\
The ranking of nodes provides interesting avenues for further research for the case where the defender flips a subset of nodes (rather than the entire network). It can be advantageous to flip the nodes with a higher PageRank more often than the other nodes. A higher PageRank means that the node has a higher probability of being compromised by a worm.

### 2015-08-19 doc/template/chap-5.tex
finaal

Another addition of Laszka et al. \cite{MitigationCovert}  to the game of FlipIt 
is extending the game to also consider non-targeted attacks by non-strategic players. In this game the defender tries to maintain control over the resource that is subjected to both targeted and non-targeted attacks. Non-targeted attacks can include phishing, while targeted attacks may include threats delivered through zero day attack vulnerabilities. \\
One of the last important additions from Laszka et al. \cite{MitigationNonTargeted} is to consider a game with targeted and non-targeted attacks where the moves made by the attacker do not succeed immediately. This approach is similar to what will be done in this paper, but nevertheless has some major differences. Firstly, in Laszka's paper, the moves by the attacker are still covert but the moves made by the defender are known to the attacker. This means that the attacker knows when the defender plays and can change its strategy depending on the moves of the defender. Our motivation for a defender with stealthy moves is that it can not be assumed that every attacker can receive feedback from an APT. Some ATPs are designed only to cause harm on the systems that are infected e.g. the Stuxnetworm that was developed to destroy nuclear reactors. The Stuxnetworm was resident on a isolated network, meaning that there was no way to connect to the internet. Another example is the use of honeypots. Honeypots emulate services or create multiple instances of real operating systems and can pretend to have sensitive information.  Honeypots do not detect all malicious attacks so the attacker can stay unnoticed and can still provide information as feedback. By providing false information through honeypots, the defender can also remain stealthy. The second difference is that even though both the targeted and non-targeted attacks do not succeed immediately, the delay is determined differently. For the targeted attack the time till it succeeds is given by an exponential distributed random variable with a known rate. The non-targeted attacks are modelled as a single attacker and the time until it succeeds is given by a Poisson process. In this paper the delay is given by one parameter, which can be the result of any virus propagation model. The third and last difference is that the paper of Laszka has multiple attackers who try to find the best strategy of the defender against both targeted and non-targeted attacks. The conclusion of Laszka's paper is that the optimal strategy for the defender is moving periodically. \\

>>>>

Another addition of Laszka et al. \cite{MitigationCovert}  to the game of FlipIt 
is extending the game to also consider non-targeted attacks by non-strategic players. In this game the defender tries to maintain control over the resource that is subjected to both targeted and non-targeted attacks. Non-targeted attacks can include phishing, while targeted attacks may include threats delivered through zero day attack vulnerabilities. \\
One of the last important additions from Laszka et al. \cite{MitigationNonTargeted} is to consider a game with targeted and non-targeted attacks where the moves made by the attacker do not succeed immediately. This approach is similar to what will be done in this paper, but nevertheless has some major differences. Firstly, in Laszka's paper, the moves by the attacker are still covert but the moves made by the defender are known to the attacker. This means that the attacker knows when the defender plays and can change its strategy depending on the moves of the defender. Our motivation for a defender with stealthy moves is that it can not be assumed that every attacker can receive feedback from an APT. Some ATPs are designed only to cause harm on the systems that are infected e.g. the \textit{Stuxnetworm} that was developed to destroy nuclear reactors. The \textit{Stuxnetworm} was resident on a isolated network, meaning that there was no way to connect to the internet. Another example is the use of honeypots. Honeypots emulate services or create multiple instances of real operating systems and can pretend to have sensitive information.  Honeypots do not detect all malicious attacks so the attacker can stay unnoticed and can still provide information as feedback. By providing false information through honeypots, the defender can also remain stealthy. The second difference is that even though both the targeted and non-targeted attacks do not succeed immediately, the delay is determined differently. For the targeted attack the time till it succeeds is given by an exponential distributed random variable with a known rate. The non-targeted attacks are modelled as a single attacker and the time until it succeeds is given by a Poisson process. In this paper the delay is given by one parameter, which can be the result of any virus propagation model. The third and last difference is that the paper of Laszka has multiple attackers who try to find the best strategy of the defender against both targeted and non-targeted attacks. The conclusion of Laszka's paper is that the optimal strategy for the defender is moving periodically. \\

### 2015-08-19 doc/template/chap-9.tex
finaal

The adaptation of the FlipIt model starts from the assumption that, when an attacker attacks at time \textit{t}, he doesn't get immediate control over the resource, but he only gains control at time \textit{t + d}, with $d$ denoting the time needed to infect a sufficient number of (or all) nodes. If the defender flips the network before the period $d$ has elapsed (so, somewhere between $t$ and $t + d$), then the attacker will never gain full control over the resource. (See figure \ref{dt} at points 3 and 4). This implies that the mathematical formulas for gain and benefit need to be adapted to the fact that the attacker loses part of his benefit because of this delay. In the remainder of this paper, we will adapt the formalization of the FlipIt game using the delay variable $d$. \\ 
We adapt the same definition for players that play at the same time as in the original FlipIt game. The probability that this happens will be close to zero, but if it happens the moves cancel each other out and no change of control happens (see figure \ref{dt} point 3).\\
\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{Images/Delayuitgelegd.pdf}
\caption{The first game is the basic FlipIt game. The second is the FlipIt game with a delay. During the first flip of the attacker, the defender moves after the delay (point 2), resulting in the attacker gaining control over the resource, until the defender flips. During the second flip of the attacker, the defender flips at time t+d, causing the defender to retain control over the resource before the attacker can gain control. During the third and final flip of the attacker, the defender flips during time t+d, causing the attacker to never gain control over the resource. }
\label{dt}
\end{figure}

%
We split the formalization in three cases. Similarly as in the original FlipIt game, there is the case where the defender plays at least as fast as the attacker, and the case where the attacker plays at least as fast as the defender. For each of these cases, the benefit formula of the basic case without delay is presented first, and subsequently the delay is introduced.  The third case is a special case where the period of the defender is smaller than the delay.\\

@@

We calculated the benefit for three cases: when $\delta_{D}$ is smaller than $d$, when $\delta_{D}$ is smaller than $\delta_{A}$ and when $\delta_{D}$ is larger than $\delta_{A}$.
The benefit formulas of case 1 and case 2.b result in the same formula. Consequently, these two formulas can be brought together by adapting the border conditions. The first figure in figure \ref{Borders} shows all the cases and it can be seen that the border condition of case 1 and case 2.b can be brought together. This can be seen as the second figure in \ref{Borders}. \\

The benefit functions can be represented as piecewise functions for each case, with case 1 and case 2.b together:\\
The benefit formula for the defender, see figure \ref{pfb} :
\begin{displaymath}
\beta_{D}(\delta_{D},\delta_{A})= \left\{
\begin{array}{lll}
	1 - \dfrac{k_{D}}{\delta_{D}}, &~~~~~ & \delta_{D} \leq d \\
	~ & ~ \\
   1 - \dfrac{(\delta_{D} - d)^2}{2\delta_{D}\delta_{A}} - \dfrac{k_{D}}{\delta_{D}}, & ~~~~~~& d \leq \delta_{D} \leq d + \delta_{A} \\
   ~ & ~ \\
   \dfrac{\delta_{A}}{2\delta_{D}} + \dfrac{d}{\delta_{D}} - \dfrac{k_{D}}{\delta_{D}} &~~~~~& \delta_{D} \geq d + \delta_{A} 
\end{array}
\right.
\end{displaymath}
~~ \\

@@

The benefit formula for the attacker, see figure \ref{pfa}:
\begin{displaymath}
\beta_{A}(\delta_{D},\delta_{A})= \left\{
\begin{array}{lll}
	 -\dfrac{k_{A}}{\delta_{A}}, & ~~~~~& \delta_{D} \leq d \\
	~ & ~ \\
   \dfrac{(\delta_{D} - d)^2}{2\delta_{D}\delta_{A}} - \dfrac{k_{A}}{\delta_{A}}, & ~~~~~~& d \leq \delta_{D} \leq d + \delta_{A} \\
   ~ & ~ \\
   1 - \dfrac{\delta_{A}}{2\delta_{D}} - \dfrac{d}{\delta_{D}} - \dfrac{k_{A}}{\delta_{A}} &~~~~~& \delta_{D} \geq d + \delta_{A} 
\end{array}
\right.
\end{displaymath}

\begin{figure}[hbtp]
\centering
\includegraphics[scale=1.4]{Images/benfa.pdf} 
\caption{The benefit function of the attacker for a cost $k_{A} =1.2$, $d =0.5$ and $\delta_{D}=5$. The three pieces match with the three colours.}
\label{pfa}
\end{figure}

The piecewise functions are also continuous functions. Filling in the border cases results in the same formula.  
\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{Images/borders.pdf}
\caption{The first figure is the representation of the cases where the benefit functions are calculated. The second figure is the representation of the new border conditions of piece B which is a merge of case 2.b and case 1.}
\label{Borders}
\end{figure}

>>>>

The adaptation of the FlipIt model starts from the assumption that, when an attacker attacks at time \textit{t}, he doesn't get immediate control over the resource, but he only gains control at time \textit{t + d}, with $d$ denoting the time needed to infect a sufficient number of (or all) nodes. If the defender flips the network before the period $d$ has elapsed (so, somewhere between $t$ and $t + d$), then the attacker will never gain full control over the resource. (See figure \ref{dt} at points 3 and 4). This implies that the mathematical formulas for gain and benefit need to be adapted to the fact that the attacker loses part of his benefit because of this delay. In the remainder of this paper, we will adapt the formalization of the FlipIt game using the delay variable $d$. \\

We apply the same definition for players that play at the same time as in the original FlipIt game. As the time in the game is continuous, the probability of this happening is close to zero, but if it happens the moves cancel each other out and no change of control happens (see figure \ref{dt} point 3).\\
\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{Images/Delayuitgelegd.pdf}
\caption{The first game is the basic FlipIt game. The second is the FlipIt game with a delay. During the first flip of the attacker, the defender moves after the delay (point 2), resulting in the attacker gaining control over the resource, until the defender flips. During the second flip of the attacker, the defender flips at time t+d, causing the defender to retain control over the resource before the attacker can gain control. During the third and final flip of the attacker, the defender flips during time t+d, causing the attacker to never gain control over the resource. }
\label{dt}
\end{figure}

%
The formalization is split three cases. Similarly as in the original FlipIt game, there is the case where the defender plays at least as fast as the attacker, and the case where the attacker plays at least as fast as the defender. For each of these cases, the benefit formula of the basic case without delay is presented first, and subsequently the delay is introduced.  The third case is a special case where the period of the defender is smaller than the delay.\\

@@

\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{Images/borders.pdf}
\caption{The first figure is the representation of the cases where the benefit functions are calculated. The second figure is the representation of the new domain of piece B which is a merge of case 2.b and case 1.}
\label{Borders}
\end{figure}
We calculated the benefit for three cases: when $\delta_{D}$ is smaller or equal to $d$, when $\delta_{D}$ is smaller or equal to $\delta_{A}$ and when $\delta_{D}$ is larger or equal to $\delta_{A}$.
The benefit formulas of case 1 and case 2.b result in the same formula. Consequently, these two formulas can be brought together by adapting the border conditions. The first figure in figure \ref{Borders} shows all the cases and it can be seen that the domain of case 1 and case 2.b can be brought together. This can be seen as the second figure in \ref{Borders}. \\

The benefit functions can be represented as piecewise functions for each case, with case 1 and case 2.b together:\\
The benefit formula for the defender is as follows:
\begin{displaymath}\label{PWFD}
\beta_{D}(\delta_{D},\delta_{A})= \left\{
\begin{array}{lll}
	1 - \dfrac{k_{D}}{\delta_{D}}, &~~~~~ & \delta_{D} \leq d \\
	~ & ~ \\
   1 - \dfrac{(\delta_{D} - d)^2}{2\delta_{D}\delta_{A}} - \dfrac{k_{D}}{\delta_{D}}, & ~~~~~~& d \leq \delta_{D} \leq d + \delta_{A} \\
   ~ & ~ \\
   \dfrac{\delta_{A}}{2\delta_{D}} + \dfrac{d}{\delta_{D}} - \dfrac{k_{D}}{\delta_{D}} &~~~~~& \delta_{D} \geq d + \delta_{A} 
\end{array}
\right.
\end{displaymath}
~~ \\

@@

The benefit formula for the attacker is as follows: 
\begin{displaymath}
\beta_{A}(\delta_{D},\delta_{A})= \left\{
\begin{array}{lll}
	 -\dfrac{k_{A}}{\delta_{A}}, & ~~~~~& \delta_{D} \leq d \\
	~ & ~ \\
   \dfrac{(\delta_{D} - d)^2}{2\delta_{D}\delta_{A}} - \dfrac{k_{A}}{\delta_{A}}, & ~~~~~~& d \leq \delta_{D} \leq d + \delta_{A} \\
   ~ & ~ \\
   1 - \dfrac{\delta_{A}}{2\delta_{D}} - \dfrac{d}{\delta_{D}} - \dfrac{k_{A}}{\delta_{A}} &~~~~~& \delta_{D} \geq d + \delta_{A} 
\end{array}
\right.
\end{displaymath}

To be able to show the continuity of the benefit function, piece 2 and piece 3 are switched and the boder conditions are reformulated.

\begin{displaymath}\label{PWFA}
\beta_{D}(\delta_{D},\delta_{A})= \left\{
\begin{array}{lll}
	1 - \dfrac{k_{D}}{\delta_{D}}, &~~~~~ & \delta_{D} \leq d \\
	~ & ~ \\
	\dfrac{\delta_{A}}{2\delta_{D}} + \dfrac{d}{\delta_{D}} - \dfrac{k_{D}}{\delta_{D}} &~~~~~& \delta_{A}  \leq \delta_{D} -d \\
	   ~ & ~ \\
   1 - \dfrac{(\delta_{D} - d)^2}{2\delta_{D}\delta_{A}} - \dfrac{k_{D}}{\delta_{D}}, & ~~~~~~&  \delta_{D} -d \leq \delta_{A}

\end{array}
\right.
\end{displaymath}
~~ \\

\begin{figure}[hbtp]
\centering
\includegraphics[scale=1.4]{Images/att.pdf} 
\caption{The benefit function of the attacker for a cost $k_{A} =0.5$, $d =3$ and $\delta_{D}=7$.}
\label{pfa}
\end{figure}

The piecewise functions are also continuous functions. Filling in the border cases results in the same formula.  This can be seen on both figures \ref{pfb} and \ref{pfa} that give respectively a visual representation of the benefit function of the defender and the benefit function of the attacker. The thick pieces indicate the cases for which each piece of the benefit function holds. For the benefit of the attacker case B and case C correspond respectively with piece 2 and piece 3 of the reformulated benefit function of the attacker.

### 2015-08-19 doc/template/conclusion.tex
finaal

\section{General Results and Conclusions}
Our FlipIt model with worm propagation delay showed us some interesting results. 
\begin{itemize}
\item When the defender plays faster than the delay, the attacker will either have a negative benefit or a benefit equal to zero if the flips do not cost anything. It is for the defender a target to be able to play at a rate smaller or equal to the delay.
\item If the defender can play with a cost equal to zero, the attacker will not play. The defender can play at any rate he wants, and the attacker is always disadvantaged by his delay.
\item If the cost of the defender is non-zero, then, depending on the ratio between it's cost and the delay, from a certain value for the speed of the attacker, the defender will not play. The same thing goes for the attacker. This result is similar to the results obtained for the original FlipIt game.
\end{itemize}

>>>>

\section{General Results and Conclusions}
Our FlipIt model with worm propagation delay showed us some interesting results. 
\begin{itemize}
\item When the defender plays faster than the delay, the attacker will either have a negative benefit or a benefit equal to zero if the flips do not cost anything. It is for the defender a target to be able to play at a rate smaller or equal to the delay.
\item If the defender can play with a cost equal to zero, the attacker will not play. The defender can play at any rate he wants, and the attacker is always disadvantaged by his delay.
\item If the cost of the defender is non-zero, then, depending on the ratio between it's cost and the delay, from a certain value for the speed of the attacker, the defender will not play. The same thing goes for the attacker. This result is similar to the results obtained for the original FlipIt game.
\item The optimum strategy for the defender and the attacker is depending on the value of \textit{d}. In the beginning a non-adaptive strategy was assumed for both players, so no player gets feedback in the game. None of the players have knowledge about the real value of \textit{d}, so both players have to make a guess. By using the model matrix based model, different kind of attacks can be simulated and an average delay can be used. 
\end{itemize}
